Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-10 paper/paper.tex
Updated

Note that  when a reachability-based  collector visits a cell  for the
first time,  it is  marked.  Explorations starting  from the  cell are
curtailed if the cell is visited subsequently due to sharing.  Since a
liveness-based collector may reach the  same cell multiple times, each
time with a different automaton state, this curtailment cannot be done
in the case  of LGC. This is an inherent  drawback of a liveness-based
collector.  This  can be mitigated  by maintaining a list  of liveness
automata states  that a  particular heap cell  was traversed  with and
avoid repeated traversals if we reach the cell in one of these states.

\subsection{Liveness-based collector with reachability-based closure
  collection}
\begin{algorithm}[t]
  \Fn{\Lgc{}}
     {
       \For {each reference $\mathit{ref}$ in root set}
            {\LCopy($\mathit{ref}$)\;}
            $\mathit{copyReferencesOnPrintStack()}$\;  
     }
     \Fn{\LCopy{$\mathit{ref}$}}
        {
          \eIf{cellType$(\mathit{ref})$ is $cons$}
              {
                \eIf {$\mathit{ref}$ is found live using dfa and has not been copied using
                  reachability}
                     {
                       $\mathit{newRef}$ =  $\mathit{dupHeapcell(ref})$\; 
                       $\mathit{ newCar}  =$  \LCopy$\mathit{(ref.car)}$\;
                       $\mathit{ newRef.car}   = \mathit{ newCar}$\;
                       $\mathit{ newCdr}  =$  \LCopy$\mathit{(ref.cdr)}$\;
                       $\mathit{ newRef.cdr}   = \mathit{newCdr}$\;
                     }
                     { $\mathit{makenull(ref)}$\;}}
              {
                \If {cellType$\mathit{(ref)}$ is closure and 
                  has not been copied}
                    { 
                      $\mathit{newRef}$  = \RCopy$(\mathit{ref})$\;
                    }
              }
        }
        \caption{LGC with reachability-based closure collection.\label{algo:lgc-b}}
\end{algorithm}
  Since the first  algorithm has  significant bookkeeping
overheads,  we have  a  second GC  scheme  that not  use liveness  for
copying closures.  In this  scheme, the liveness automata is consulted
only for \CONS\ cells; we assume everything under a closure to be live
and copy everything  that is reachable. Algorithm~\ref{algo:lgc-b}  describes  this scheme.

A consequence of copying everything under a closure is that we can end
up trying to copy things which are not live and corrupt the heap. This
can happen if a \CONS\  cell copied using liveness during a collection
subsequently  becomes   part  of  a  closure.    During  a  subsequent
collection,  the  collector  may  try  to copy  everything  under  the
\CONS\ cell thus copying heap  locations which might not be valid.  To
avoid  this, we  should ensure  that any  non-live references  are not
carried across  GC's.  We  do this  by scanning the  live part  of the
buffer after every  GC and setting to null any  reference which is not
live. This  is correct  because any reference  which is dead  during a
collection will not be dereferenced any further.

Using  reachablity  for  closures   avoids  the  problem  of  multiple
explorations from the  same cell.  This is done by  maintaining a flag
in each cell to indicate whether it was copied using reachability.

>>>>

Note that  when a reachability-based  collector visits a cell  for the
first time,  it is  marked.  Explorations starting  from the  cell are
curtailed if the cell is visited subsequently due to sharing.  Since a
liveness-based collector may reach the  same cell multiple times, each
time with a different automaton state, this curtailment cannot be done
in the case  of LGC. This is an inherent  drawback of a liveness-based
collector.  This  can be mitigated  by maintaining a list  of liveness
automata states  that a  particular heap cell  was traversed  with and
avoid repeated traversals if we reach the cell in one of these states.
However this also involves storage and run-time penalty.

To  avoid  performance  penalty,  we experimented  with  a  mixed-mode
garbage collection scheme  which uses liveness for  evaluated data and
reachability for closures.

\subsection{Liveness-based collector with reachability-based closure
  collection}
\begin{algorithm}[t]
  \Fn{\Lgc{}}
     {
       \For {each reference $\mathit{ref}$ in root set}
            {\LCopy($\mathit{ref}$)\;}
            $\mathit{copyReferencesOnPrintStack()}$\;  
     }
     \Fn{\LCopy{$\mathit{ref}$}}
        {
          \eIf{cellType$(\mathit{ref})$ is $cons$}
              {
                \eIf {$\mathit{ref}$ is found live using dfa and has not been copied using
                  reachability}
                     {
                       $\mathit{newRef}$ =  $\mathit{dupHeapcell(ref})$\; 
                       $\mathit{ newCar}  =$  \LCopy$\mathit{(ref.car)}$\;
                       $\mathit{ newRef.car}   = \mathit{ newCar}$\;
                       $\mathit{ newCdr}  =$  \LCopy$\mathit{(ref.cdr)}$\;
                       $\mathit{ newRef.cdr}   = \mathit{newCdr}$\;
                     }
                     { $\mathit{makenull(ref)}$\;}}
              {
                \If {cellType$\mathit{(ref)}$ is closure and 
                  has not been copied}
                    { 
                      $\mathit{newRef}$  = \RCopy$(\mathit{ref})$\;
                    }
              }
        }
        \caption{LGC with reachability-based closure collection.\label{algo:lgc-b}}
\end{algorithm}
  Algorithm~\ref{algo:lgc-b}  describes  a garbage  collection  scheme
  where the liveness automata is  consulted only for \CONS\ cells; and
  assumes everything under a closure  to be live and copies everything
  that is reachable.

A consequence of  copying everything under a closure is  that we might
reach  a non-live  cell that  has been  reclaimed earlier  by liveness
based GC.   This can  happen if  a \CONS\  cell having  partially live
sub-structure  is copied  using liveness  during a  collection and  it
subsequently  becomes part  of a  live closure.   During a  subsequent
collection, the collector may try to copy everything under the closure
including the dead  sub-structure of the \CONS\ cell.   To avoid this,
we ensure  that any non-live  references are not carried  across GC's.
We do this by scanning the live  part of the buffer after every GC and
nullifying any  reference which is  not live. This is  correct because
any reference which  is dead during a collection is  guaranteed not be
dereferenced any further by the program.

This mixed scheme avoids the problem of multiple explorations from the
same  cell.  This  is  done by  maintaining  a flag  in  each cell  to
indicate whether it was copied using reachability.
